A general and efficient representation of Ancestral Recombination Graphs

### 2023-09-14 paper.tex

The word ``ARG'' is now widely used to refer to genetic genealogies in general~\citep[MORE][]{mathieson2020ancestry}, including
varied outputs from modern simulation and
inference methods~\citep{rasmussen2014genome, palamara2016argon, haller2018tree, kelleher2019inferring, speidel2019method, baumdicker2021efficient, zhang2023biobank}.
However, this broad usage means there is disagreement on the exact object being defined. This causes problems, for example,
not just when evaluating inferred ARGs, but even when defining
what the criteria for success should be.
It also means there is no shared format for interchanging ARG data between
programs, resulting in the assorted negative consequences with which
geneticists are only too familiar~\citep{excoffier2006computer}.
Perhaps most importantly some definitions of ARGs, especially those linked to a specific stochastic process, are overly narrow which
can lead to missed opportunities in their application and usage.
Such problems go much deeper than pedantic discussion over terminology;
they directly inform what we should be attempting to estimate, how we
assess success, and the expected limitations on the precision
and accuracy of our inferences.

@@

A gARG is then a graph in which nodes represent
haploid genomes, edges represent
genetic inheritance between an ancestor and a descendant, and
annotations specify coordinates of the genomic intervals
over which genetic inheritance occurs.
We can therefore view a gARG for a given set of individuals
as a subset of their pedigree.
An example of this for diploids (so each individual is represented by two nodes) is illustrated in Fig.~\ref{fig-arg-in-pedigree}.
Each genome is inherited from one
of the individual's parents,
and may be the recombined product of that parent's two genomes.
For example, individual $I_1$ in Fig.~\ref{fig-arg-in-pedigree}
has two genomes \textsf{a} and \textsf{b},
inherited from parents $I_3$ and $I_4$. Genome \textsf{a} is the product of
recombining $I_3$'s two genomes \textsf{e} and \textsf{f} at position 2,
whereas \textsf{b} was inherited directly from $I_4$'s genome \textsf{g} without
recombination.

@@

The topology of a gARG specifies that genetic inheritance occurred between particular
ancestors and descendants, but the graph connectivity
does not tell us which \emph{parts} of their genomes were inherited.
In order to capture the effects of recombination (and any other process
that results in the partial inheritance of a descendant genome
from an ancestor) we ``annotate'' the edges with the genome
coordinates over which inheritance occurs.
Continuing the example of genome \textsf{a} in Fig.~\ref{fig-arg-in-pedigree},
the recombination at position 2 means that it inherits the (half-closed) genome
interval $[0, 2)$ from \textsf{e} and it inherits $[2, 10)$ from \textsf{f}; these intervals are shown attached to the corresponding graph edges. Alternative gARG encodings split transmitted intervals and associate each with a new node, resulting in a single genome being represented by a group of multiple nodes (or ``blocks'' \citep{palamara2016argon}), with implications for efficiency of storage.

% JK: dunno if this will make some readers switch off, but I feel it's
% important to show that this is a formal model and the concrete definition
% will help some readers?
Formally, our gARG encoding is defined by sets of nodes $N$ and edges $E$, where each
element of $E$ is a tuple $(c, p, I)$ such that $c, p \in N$ are the child and
parent nodes and $I$ is the set of disjoint genomic intervals $[\ell, r)$ over
which genome $c$ inherits from $p$. The most fundamental operation for
an ARG data structure is the process of recovering the corresponding sequence
of local trees along the genome. Concretely, we can describe a local tree
at position $x$ as an oriented forest $\pi = ( \pi_1, \hdots, \pi_{|N|} )$, where the entry $\pi_j$ is the index
of the parent of $j$ and $\pi_j = \emptyset$ if $j$ is not present in the
tree~\citep{kelleher2013coalescent,kelleher2014coalescent,kelleher2016efficient}.
We use a forest to describe the local ancestry at a given position so that we allow for cases where not all leaves have found a most recent common ancestor, resulting in multiple roots.
To construct $\pi$ at position $x$, set $\pi_i = \emptyset$ for $i \in \{1, \hdots, |N|\}$.
Then, traversing over all nodes of the tree,
at each node
$u$, find an edge $(c, p, I) \in E$ such that $u = c$ and $x \in I$, and set
$\pi_c \leftarrow p$.
At the end of this process, $\pi$ fully describes the local tree.

@@

@@

Storing only breakpoint positions, rather than transmitted regions,
makes it hard to perform genetic analyses (such as extracting local trees)
without traversing the entire graph.
% "For this reason"?
Some eARG encodings, such as \citet{shipilina2023origin}, have a specialised definition of edges, which places them outside the scope of the standard graph-theoretic representation.
%% Is the above true?? It the reason for this to allow efficient tree extraction? Are there other examples where encodings have been
%% built on top of an eARG-like node annotation scheme (ARGON?) and made efficient somehow?

@@

\section*{From eARGs to sample-resolved gARGs}
Any eARG can be converted to a gARG without loss of information,
and the conversion process is very similar to
simplifying a prospective gARG with respect to a set of samples.
The process involves two steps, and is illustrated in
Fig.~\ref{fig-ancestry-resolution}.
The first step is straightforward
(Fig.~\ref{fig-ancestry-resolution}B).
We duplicate the nodes and
edges from the input eARG, and the add inheritance annotations
to the output gARG's edges. If the node is a common ancestor node,
we annotate the single outbound edge with the interval $[0,L)$,
for a genome of length $L$.
If the node is a recombination event with a breakpoint $x$,
we annotate the outbound edges with the intervals $[0, x)$
and $[x, L)$.
These inheritance annotations are clearly in one-to-one
correspondance with the information in the input eARG,
but they are a superset of the
``ancestral material''~\citep{wiuf1999ancestry,wiuf1999recombination}
of the sample.

@@

\section*{Levels of simplification}\label{ARG_simplification}
ARG simplification is a powerful tool, and can do much more
besides removing unreachable nodes and edges in a prospective ARG
and resolving ancestral material in an eARG.
In general, we can think of
ARG simplification as the process
of removing nodes and re-writing edges (and their inheritance annotations)
to remove various types of redundancy.
The redundancy that we are interested in
revolves around the presence
of ``unary'' regions for nodes
(where a node has exactly one child over a particular genomic region). These arise in several different ways.
Nodes with one child are not a standard feature of evolutionary trees;
we are usually interested in
nodes that have at least two children, because these
are in principle detectable from mutational differences
arising in the respective subtrees.
A tree branch may correspond to many generations
of individuals (or indeed cell divisions) and there is usually
very little information about these intermediates, or utility
in modelling them.
% FIXME the bridge between the next few paragraphs is a bit weird, in
% that we've just been talking about how interesting unary nodes are
% but then go on about how to remove them. Just needs a bit of smoothing
% out I think.
ARGs, however, are different, and there
can be significant information about evolutionary processes
provided by nodes which appear unary in one or more local trees.

\begin{figure}
\centering
\vspace{5em}
\includegraphics[width=\linewidth]{illustrations/simplification}
\caption{\label{fig-simplification}
Levels of ARG simplification.
In all cases the graph visualization is shown on the left
and the corresponding local trees on the right.
Nodes are coloured by the number of parents and shaded
according to the proportion of their span over which they are coalescent.
(A) A gARG simulated from a backward-time diploid Wright-Fisher
model: genome \noderef{k} has three children, and the genomes
\noderef{e} and \noderef{g} simultaneously have multiple parents and multiple children.
(B) Simplified to remove all
singly-connected graph components (e.g., diamonds such as \noderef{jlnm}).
(C) Remove nodes that never represent coalescences, i.e. are unary everywhere (e.g.\ \noderef{n} and \noderef{r}).
(D) Rewrite edges to bypass any unary nodes in local trees.
}
\end{figure}

Fig.~\ref{fig-simplification} show a series of successive steps of
simplification, starting with a sample-resolved but unsimplified gARG, simulated under a
backward-time Wright-Fisher
model. We show the graph on the left, and the corresponding sequence of
local trees on the right.
In Fig.~\ref{fig-simplification}A the local trees (right)
contain many unary nodes, and as we successively simplify
the local trees contain fewer
unary nodes (Fig.~\ref{fig-simplification}B,C) until
we reach Fig.~\ref{fig-simplification}D, where the local trees
have no unary nodes.
A node is unary in a local tree covering the genome interval $I$
if genetic inheritance passes through that ARG node,
but no coalescence occurs in the interval $I$.
The distinction between the ``common ancestry'' of two or more genomes
in an ancestral genome and the ``coalescence'' which may or may
not occur in the local trees is
important~\citep{hudson1983testing,kelleher2016efficient}.
Consider \noderef{e} in in Fig.~\ref{fig-simplification}A,
for example. We can see from the graph that it is a common
ancestor of  samples \noderef{a} and \noderef{b}, but
it does not correspond to any coalescence in the
local trees to the left of position $44$, and is a therefore a
unary node in these three trees.

>>>>

The word ``ARG'' is now widely used to refer to genetic genealogies in
general~\citep[MORE][]{mathieson2020ancestry}, including
varied outputs from modern simulation and
inference methods~\citep{rasmussen2014genome, palamara2016argon, haller2018tree,
kelleher2019inferring, speidel2019method, baumdicker2021efficient, zhang2023biobank}.
However, this broad usage means there is disagreement on the
exact object being defined. This causes problems, for example,
not just when evaluating inferred ARGs, but even when defining
what the criteria for success should be.
It also means there is no shared format for interchanging ARG data between
programs, resulting in the assorted negative consequences with which
geneticists are only too familiar~\citep{excoffier2006computer}.
Perhaps most importantly, some definitions of ARGs
(especially those linked to a specific stochastic process) are overly narrow, which
can lead to missed opportunities in their application and usage.
Such problems go much deeper than pedantic discussion over terminology;
they directly inform what we should be attempting to estimate, how we
assess success, and the expected limitations on the precision
and accuracy of our inferences.

@@

A gARG is then a graph in which nodes represent
haploid genomes, edges represent
genetic inheritance between an ancestor and a descendant, and
annotations specify coordinates of the genomic intervals
over which genetic inheritance occurs.
% FIXME WE're missing the important distinction here that the pedigree
% has one node per ind, whereas a gARG has two nodes.
We can therefore view a gARG for a given set of individuals
as a subset of their pedigree.
An example of this for diploids
(so each individual is represented by two nodes)
is illustrated in Fig.~\ref{fig-arg-in-pedigree}.
Each genome is inherited from one
of the individual's parents,
and may be the recombined product of that parent's two genomes.
For example, individual $I_1$ in Fig.~\ref{fig-arg-in-pedigree}
has two genomes \textsf{a} and \textsf{b},
inherited from parents $I_3$ and $I_4$. Genome \textsf{a} is the product of
recombining $I_3$'s two genomes \textsf{e} and \textsf{f} at position 2,
whereas \textsf{b} was inherited directly from $I_4$'s genome \textsf{g} without
recombination.

@@

The topology of a gARG specifies that genetic inheritance occurred between particular
ancestors and descendants, but the graph connectivity
does not tell us which \emph{parts} of their genomes were inherited.
In order to capture the effects of recombination (and any other process
that results in the partial inheritance of a descendant genome
from an ancestor) we ``annotate'' the edges with the genome
coordinates over which inheritance occurs.
Continuing the example of genome \textsf{a} in Fig.~\ref{fig-arg-in-pedigree},
the recombination at position 2 means that it inherits the (half-closed) genome
interval $[0, 2)$ from \textsf{e} and it inherits $[2, 10)$ from \textsf{f};
these intervals are shown attached to the corresponding graph edges.

% important to show that this is a formal model and the concrete definition
% will help some readers?
% FIXME - the oriented forest definition is wrong, we need to be more
% concrete about the fact that nodes are integers.
Formally, we define a gARG as a sets of nodes $N$ and edges $E$.
Each element of $E$ is a tuple $(c, p, I)$ such that $c, p \in N$ are the child and
parent nodes and $I$ is the set of disjoint genomic intervals $[\ell, r)$ over
which genome $c$ inherits from $p$. The most fundamental operation for
an ARG data structure is the process of recovering the corresponding sequence
of local trees along the genome. Concretely, we can describe a local tree
at position $x$ as an oriented forest $\pi = ( \pi_1, \hdots, \pi_{|N|} )$,
where the entry $\pi_j$ is the index
of the parent of $j$ and $\pi_j = \emptyset$ if $j$ is not present in the
tree~\citep{kelleher2013coalescent,kelleher2014coalescent,kelleher2016efficient}.
We use a forest to describe the local ancestry at a given
position so that we allow for cases where not all leaves have
found a most recent common ancestor, resulting in multiple roots.
To construct $\pi$ at position $x$, set
$\pi_i = \emptyset$ for $i \in \{1, \hdots, |N|\}$.
% FIXME - this is wrong
Then, traversing over all nodes of the tree,
at each node
$u$, find an edge $(c, p, I) \in E$ such that $u = c$ and $x \in I$, and set
$\pi_c \leftarrow p$.
At the end of this process, $\pi$ fully describes the local tree.

@@

[FIXME find somewhere natural for this. The last clause is too vague - what
are the implications?]
Alternative gARG encodings split transmitted intervals and associate each with
a new node, resulting in a single genome being represented by a group of
multiple nodes (or ``blocks'' \citep{palamara2016argon}), with implications for
efficiency of storage.

@@

% FIXME this isn't worth getting into
Storing only breakpoint positions, rather than transmitted regions,
makes it hard to perform genetic analyses (such as extracting local trees)
without traversing the entire graph.
% "For this reason"?
%% Is the above true?? It the reason for this to allow efficient tree extraction? Are there other examples where encodings have been
%% built on top of an eARG-like node annotation scheme (ARGON?) and made efficient somehow?

@@

% JK: previous heading was jargony - this is clearer
\section*{Converting an eARG to a gARG}
Any eARG can be converted to a gARG without loss of information,
and the conversion process is very similar to
simplifying a prospective gARG with respect to a set of samples.
The process involves two steps, and is illustrated in
Fig.~\ref{fig-ancestry-resolution}.
The first step is straightforward
(Fig.~\ref{fig-ancestry-resolution}B).
We duplicate the nodes and
edges from the input eARG, and the add inheritance annotations
to the output gARG's edges. If the node is a common ancestor node,
we annotate the single outbound edge with the interval $[0,L)$,
for a genome of length $L$.
If the node is a recombination event with a breakpoint $x$,
we annotate the outbound edges with the intervals $[0, x)$
and $[x, L)$.
These inheritance annotations are clearly in one-to-one
correspondance with the information in the input eARG,
but they are a superset of the
``ancestral material''~\citep{wiuf1999ancestry,wiuf1999recombination}
of the sample.

@@

\section*{Levels of simplification}\label{ARG_simplification}
ARG simplification is a powerful tool, and can do much more
besides removing unreachable nodes and edges in a prospective ARG
and resolving ancestral material in an eARG.
In general, we can think of
ARG simplification as the process
of removing nodes and re-writing edges (and their inheritance annotations)
to remove various types of redundancy.
The redundancy that we are interested in
revolves around the presence
``unary'' nodes in the local trees
(where a node has exactly one child over a particular genomic region).
% "unary regions" is jargon - just dropping it in here like this doesn't
% clarify. We can't assume that people are used to switching between
% graphs and trees.
% of ``unary'' regions for nodes
% (where a node has exactly one child over a particular genomic region).
These arise in several different ways.
Nodes with one child are not a standard feature of evolutionary trees;
we are usually interested in
nodes that have at least two children, because these
are in principle detectable from mutational differences
arising in the respective subtrees.
A tree branch may correspond to many generations
of individuals (or indeed cell divisions) and there is usually
very little information about these intermediates, or utility
in modelling them.
% FIXME the bridge between the next few paragraphs is a bit weird, in
% that we've just been talking about how interesting unary nodes are
% but then go on about how to remove them. Just needs a bit of smoothing
% out I think.
ARGs, however, are different, and there
can be significant information about evolutionary processes
provided by nodes that are unary in one or more local trees.

\begin{figure}
\centering
\vspace{5em}
\includegraphics[width=\linewidth]{illustrations/simplification}
\caption{\label{fig-simplification}
Levels of ARG simplification.
In all cases the graph visualization is shown on the left
and the corresponding local trees on the right.
Nodes are coloured by the number of parents and shaded
according to the proportion of their span over which they are coalescent.
(A) A gARG simulated from a backward-time diploid Wright-Fisher
model: genome \noderef{k} has three children, and the genomes
\noderef{e} and \noderef{g} simultaneously have multiple parents and multiple children.
(B) Simplified to remove all
singly-connected graph components (e.g., diamonds such as \noderef{jlnm}).
(C) Remove nodes that never represent coalescences,
i.e.\ are unary everywhere (e.g.\ \noderef{n} and \noderef{r}).
(D) Rewrite edges to bypass any unary nodes in local trees.
}
\end{figure}

Fig.~\ref{fig-simplification} shows a series of successive steps of
simplification,
% FIXME this doesn't explain anything - what did we store in the simulation?
starting with a sample-resolved but unsimplified gARG, simulated under a
backward-time Wright-Fisher
model. We show the graph on the left, and the corresponding sequence of
local trees on the right.
In Fig.~\ref{fig-simplification}A the local trees (right)
contain many unary nodes, and as we successively simplify
the local trees contain fewer
unary nodes (Fig.~\ref{fig-simplification}B,C) until
we reach Fig.~\ref{fig-simplification}D, where the local trees
have no unary nodes.
A node is unary in a local tree covering the genome interval $I$
if genetic inheritance passes through that ARG node,
but no coalescence occurs in the interval $I$.
The distinction between the ``common ancestry'' of two or more genomes
in an ancestral genome and the ``coalescence'' which may or may
not occur in the local trees is
important~\citep{hudson1983testing,kelleher2016efficient}.
Consider \noderef{e} in Fig.~\ref{fig-simplification}A,
for example. We can see from the graph that it is a common
ancestor of  samples \noderef{a} and \noderef{b}, but
it does not correspond to any coalescence in the
local trees to the left of position $44$, and is a therefore a
unary node in these three trees.
